---
title: "Measles outbreak simulation in Miami"
---

::: {.callout-note}
The original version of this document is part of the ID Cup project and can be found at <https://github.com/EpiForeSITE/idcup>. This version is adapted for the InsightNet 2026 course and may contain edits for clarity or relevance to the course content.
:::

::: {.callout-caution title="Work in progress"}
This scenario simulation is a work in progress. We are still reviewing the model, assumptions, and parameters. Please be mindful of this when interpreting the results.
:::

## Preparing the environment

```{r}
#| label: loading-libraries
#| warning: false
#| message: false
library(measles)
library(data.table)

params <- list(
  city = "Miami",
  max_pop = 10000,
  nthreads = 4,
  seed = 8812,
  ndays = 60,
  nsims = 20
  )
```

Pulling information about the city and population from the data files. The age distribution is county-level; for Miami, the data are from Miami-Dade County. The mixing matrix is based on EpiStorm [@litvinovaEpistormMixMappingSocial2025] survey data, which provides age-based contact patterns. When the data file contains multiple city matrices, we use the matrix corresponding to `params$city`; otherwise, we use the single matrix included with the workshop materials.

```{r}
#| label: loading-data
# Population matrix
census_age <- fread("data/census_age.csv")[city == params$city]
population <- fread("data/population.csv")[city == params$city, Population]

# Mixing matrix (age-based, from EpiStorm survey data)
mixing_matrices <- readRDS("data/mixing_matrix.rds")
if (is.list(mixing_matrices) && !is.matrix(mixing_matrices)) {
  if (!params$city %in% names(mixing_matrices)) {
    stop(sprintf("No mixing matrix found for %s.", params$city))
  }
  mixing_matrix <- mixing_matrices[[params$city]]
} else {
  mixing_matrix <- mixing_matrices
}

# Retrieving vaccination rate
vacc_rate <- fread("data/mmr.csv")[city == params$city, vacc_rate] / 100
```

The total population in the city data, `r prettyNum(population)`, exceeds the workshop target population size (`r prettyNum(params$max_pop)`). The age distribution will be scaled down to keep the live example small.

```{r}
#| label: adjusting-population
N <- min(params$max_pop, sum(population))

census_age[, total := ceiling(total/sum(total) * N)]
census_age[, vacc_rate := vacc_rate]

N <- sum(census_age$total)
```

```{r}
#| label: calibrating-mixing
calib_scale <- calibrate_mixing_model(
  contact_matrix = mixing_matrix,
  target_rep_number = 15,
  infectious_period_days = 4,
  transmission_prob = 0.2
  )
```

## Baseline scenario

```{r}
#| label: creating-model

# Finding the hospitalization rate for
# a 10% hospitalization probability:
# P(hosp) = hosp_r / (hosp_r + rec_r)
#   => hosp_r = P(hosp) (hosp_r + rec_r)
#             = P(hosp) * hosp_r + P(hosp) * rec_r
#   => hosp_r = P(hosp) * rec_r / (1 - P(hosp))
h_rate <- 0.1 * (1 / 3) / (1 - 0.1)

measles_model <- ModelMeaslesMixing(
  n                            = N,
  prevalence                   = 1,
  transmission_rate            = 0.2,
  vax_efficacy                 = 0.97,
  contact_matrix               = mixing_matrix * calib_scale,
  hospitalization_rate         = h_rate,
  hospitalization_period       = 7,
  # Intervention parameter
  quarantine_period            = 14,
  quarantine_willingness       = 0.9,
  isolation_willingness        = 0.9,
  isolation_period             = 4,
  # Vaccination parameters (this will be replaced)
  prop_vaccinated              = 0.95,
  # Contact tracing parameters
  days_undetected              = 2,
  contact_tracing_success_rate = 0.8,
  contact_tracing_days_window  = 4,
  rash_period                  = 3
)
```


```{r}
#| label: adding-entities
# Adding the entities to the model
add_entities_from_dataframe(
  model = measles_model,
  entities = census_age,
  col_name = "age_group",
  col_number = "total",
  as_proportion = FALSE
)
```

For `r params$city`, the vaccination rate is approximately `r sprintf("%.2f", census_age[, sum(total * vacc_rate)/sum(total)] * 100)`%.

```{r}
#| label: setting-distribution
# Creating the distribution function
dist_fun <- distribute_tool_to_entities(
  prevalence = census_age$vacc_rate,
  as_proportion = TRUE
)

# Setting the distribution function
measles_model |>
  get_tool(0) |>
  set_distribution_tool(dist_fun)
```

For the initial cases, we will use 2 active cases and distribute them randomly across the population.

```{r}
#| label: distribute-initial-cases
get_virus(measles_model, 0L) |>
  set_distribution_virus(
    distfun = distribute_virus_randomly(
      prevalence = 2,
      as_proportion = FALSE
      )
  )
```

We can now run the simulations. The workshop defaults use `r params$nsims` simulations and a maximum population of `r prettyNum(params$max_pop)` so the example can run quickly during a live session. For a fuller analysis, increase `params$nsims` and `params$max_pop`.

```{r}
#| label: running-simulations
saver <- make_saver("outbreak_size", "hospitalizations", "transition")

measles_model |>
  run_multiple(
    ndays = params$ndays,
    nsims = params$nsims,
    seed  = params$seed,
    saver = saver,
    nthreads = params$nthreads
  )
```

We can check the results of the simulations by summarizing the model:

```{r}
#| label: checking-simulations
summary(measles_model)
```

## Extracting and plotting the results

```{r}
#| label: extracting-results
# Extracting the results
ans <- measles_model |>
  run_multiple_get_results(
    freader = data.table::fread
  )

# Converting into data.table format for convenience
outbreak_size <- ans$outbreak_size |> as.data.table()
hospitalizations <- ans$hospitalizations |> as.data.table()
transition <- ans$transition |> as.data.table()
```

### Transition

```{r}
# Preprocessing
transition
# To get a starting point
tmat <- get_transition_probability(measles_model) * 0

# Collapsing the data
t_totals <- transition[, .(total = sum(counts)), by = .(from, to)]
tmat[
  as.matrix(t_totals[, .(from, to)])
  ] <- t_totals$total

draw_mermaid_from_matrix(tmat) |> 
  plot()
```

### Outbreak size

```{r}
#| label: plotting-outbreak-size1
# Plotting outbreak size over time
ggplot2::ggplot(
  outbreak_size,
  ggplot2::aes(x = date, y = outbreak_size)
) +
  ggplot2::geom_point(
    color = "blue",
    alpha = 0.2,
    size = 1.5
  ) +
  ggplot2::labs(
    title = paste("Measles outbreak size in", params$city),
    subtitle = sprintf(
      "Mixing model with census data (%i simulations)",
      params$nsims
    ),
    x = "Day",
    y = "Cases"
  ) +
  ggplot2::theme_minimal()
```


```{r}
#| label: plotting-outbreak-size2
# Aggregating the final day
final_outbreak <- outbreak_size[date == max(date)]
median_outbreak_size <- median(final_outbreak$outbreak_size)

ggplot2::ggplot(final_outbreak, ggplot2::aes(x = outbreak_size)) +
  ggplot2::geom_histogram(fill = "steelblue", color = "white") +
  ggplot2::labs(
    title = sprintf(
      "Measles outbreak size in %s (median: %.2f)",
      params$city,
      median_outbreak_size
    ),
    subtitle = sprintf(
      "Mixing model with census data (%i simulations)",
      params$nsims
    ),
    x = "Outbreak size",
    y = "Count"
  ) +
  ggplot2::theme_minimal()
```

We can save the results for further analysis.

```{r}
#| label: saving-results
fwrite(outbreak_size, sprintf("%s_outbreak_size.csv", params$city))
fwrite(hospitalizations, sprintf("%s_hospitalizations.csv", params$city))
```

## Session info

This document was generated using the `{measles}` R package version `r packageVersion("measles")` and the `{epiworldR}` package version `r packageVersion("epiworldR")` on `r Sys.Date()`. We used R version `r getRversion()`, running on `r Sys.info()[['sysname']]` with `r Sys.info()[['machine']]` architecture.

## References
