---
title: "Measles outbreak simulation in Miami"
---

::: {.callout-note}
The original version of this document is part of the ID Cup project and can be found at <https://github.com/EpiForeSITE/idcup>. This version is adapted for the InsightNet 2026 course and may contain edits for clarity or relevance to the course content.
:::

::: {.callout-caution title="Work in progress"}
This scenario simulation is a work in progress. We are still reviewing the model, assumptions, and paramters. Please be mindful of this when interpreting the results.
:::

## Preparing the environment

```{r}
#| label: loading-libraries
#| warning: false
#| message: false
library(measles)
library(data.table)

params <- list(
  city = "Miami",
  max_pop = 50000,
  nthreads = 4,
  seed = 8812,
  ndays = 60,
  nsims = 100
  )
```

Pulling information about the city and population from the data files. The mixing matrix is based on the POLYMOD survey data, which provides age-based contact patterns for various cities. We will use the mixing matrix corresponding to the specified city in `params$city`. If the city is not found in the available mixing matrices, an error will be raised.

```{r}
#| label: loading-data
# Population matrix
census_age <- fread("data/census_age.csv")[city == params$city]
population <- 487014

# Mixing matrix (age-based, from POLYMOD survey data)
mixing_matrices <- readRDS("data/mixing_matrix.rds")[[params$city]]

# Retrieving vaccination rate
vacc_rate <- 0.881
```

The total population in the census data `r population`, exceeds the maximum population size (`r params$max_pop`). The population will be scaled down to fit within the specified maximum population size.

```{r}
#| label: adjusting-population
N <- min(params$max_pop, sum(population))

census_age[, total := ceiling(total/sum(total) * N)]
census_age[, vacc_rate := vacc_rate]

N <- sum(census_age$total)
```

```{r}
#| label: calibrating-mixing
calib_scale <- calibrate_mixing_model(
  contact_matrix = mixing_matrices,
  target_rep_number = 15,
  infectious_period_days = 4,
  transmission_prob = 0.2
  )
```

## Baseline scenario

```{r}
#| label: creating-model

# Finding hospitalization rate for
# a 10% hospitalization probability
# P(hosp) = hosp_r / (hosp_r + rec_r)
#   => hosp_r = P(hosp) (hosp_r + rec_r)
#             = P(hosp) * hosp_r + P(hosp) * rec_r
#   => hosp_r = P(hosp) * rec_r / (1 - P(hosp))
h_rate <- 0.1 * (1 / 3) / (1 - 0.1)

measles_model <- ModelMeaslesMixing(
  n                            = N,
  prevalence                   = 1,
  transmission_rate            = 0.2,
  vax_efficacy                 = 0.97,
  contact_matrix               = mixing_matrices * calib_scale,
  hospitalization_rate         = h_rate,
  hospitalization_period       = 7,
  # Intervention parameter
  quarantine_period            = 14,
  quarantine_willingness       = 0.9,
  isolation_willingness        = 0.9,
  isolation_period             = 4,
  # Vaccination parameters (this will be replaced)
  prop_vaccinated              = 0.95,
  # Contact tracing parameters
  days_undetected              = 2,
  contact_tracing_success_rate = 0.8,
  contact_tracing_days_window  = 4,
  rash_period                  = 3
)
```


```{r}
#| label: adding-entities
# Adding the entities to the model
add_entities_from_dataframe(
  model = measles_model,
  entities = census_age,
  col_name = "age_group",
  col_number = "total",
  as_proportion = FALSE
)
```

For `r params$city`, the vaccination rate is approximately `r sprintf("%.2f", census_age[, sum(total * vacc_rate)/sum(total)] * 100)`%.

```{r}
#| label: setting-distribution
# Creating the distribution function
dist_fun <- distribute_tool_to_entities(
  prevalence = census_age$vacc_rate,
  as_proportion = TRUE
)

# Setting the distribution function
measles_model |>
  get_tool(0) |>
  set_distribution_tool(dist_fun)
```

For the initial cases, we will use the expected number of active cases based on the reported cases and the sampling approach described above. We will distribute these initial cases randomly across the population.

```{r}
#| label: distribute-initial-cases
get_virus(measles_model, 0L) |>
  set_distribution_virus(
    distfun = distribute_virus_randomly(
      prevalence = 2,
      as_proportion = FALSE
      )
  )
```

We will use 2 initial cases in the simulations, distributed randomly across the population. We can now run the simulations.

```{r}
#| label: running-simulations
saver <- make_saver("outbreak_size", "hospitalizations", "transition")

measles_model |>
  run_multiple(
    ndays = params$ndays,
    nsims = params$nsims,
    seed  = params$seed,
    saver = saver,
    nthreads = params$nthreads
  )
```

We can check the results of the simulations by summarizing the model:

```{r}
#| label: checking-simulations
summary(measles_model)
```

## Extracting and plotting the results

```{r}
#| label: extracting-results
# Extracting the results
ans <- measles_model |>
  run_multiple_get_results(
    freader = data.table::fread
  )

# Converting into data.table format for convenience
outbreak_size <- ans$outbreak_size |> as.data.table()
hospitalizations <- ans$hospitalizations |> as.data.table()
transition <- ans$transition |> as.data.table()
```

### Transition

```{r}
# Preprocessing
transition
# To get a starting point
tmat <- get_transition_probability(measles_model) * 0

# Collapsing the dat
t_totals <- transition[, .(total = sum(counts)), by = .(from, to)]
tmat[
  as.matrix(t_totals[, .(from, to)])
  ] <- t_totals$total

draw_mermaid_from_matrix(tmat) |> 
  plot()
```

### Outbreak size

```{r}
#| label: plotting-outbreak-size1
# Aggregating to get the final
ggplot2::ggplot(
  outbreak_size,
  ggplot2::aes(x = date, y = outbreak_size)
) +
  ggplot2::geom_point(
    color = "blue",
    alpha = 0.2,
    size = 1.5
  ) +
  ggplot2::labs(
    title = paste("Measles outbreak size in", params$city),
    subtitle = sprintf(
      "Mixing model with census data (%i simulations)",
      params$nsims
    ),
    x = "Day",
    y = "Cases"
  ) +
  ggplot2::theme_minimal()
```


```{r}
#| label: plotting-outbreak-size2
# Aggregating to get the final
final_outbreak <- outbreak_size[date == max(date)]
median_outbreak_size <- median(final_outbreak$outbreak_size)

ggplot2::ggplot(final_outbreak, ggplot2::aes(x = outbreak_size)) +
  ggplot2::geom_histogram(fill = "steelblue", color = "white") +
  ggplot2::labs(
    title = sprintf(
      "Measles outbreak size in %s (median: %.2f)",
      params$city,
      median_outbreak_size
    ),
    subtitle = sprintf(
      "Mixing model with census data (%i simulations)",
      params$nsims
    ),
    x = "Outbreak size",
    y = "Count"
  ) +
  ggplot2::theme_minimal()
```

We can save the results for further analysis.

```{r}
#| label: saving-results
fwrite(outbreak_size, sprintf("%s_outbreak_size.csv", params$city))
fwrite(hospitalizations, sprintf("%s_hospitalizations.csv", params$city))
```

## Session info

This document was generated using the `{measles}` R package version `r packageVersion("measles")` and the `{epiworldR}` package version `r packageVersion("epiworldR")` on `r Sys.Date()`. We used R version `r getRversion()`, running on `r Sys.info()[['sysname']]` with `r Sys.info()[['machine']]` architecture.
