---
title: Getting Started
---

`epiworld` is an advanced, highly flexible and performant Agent-Based Modeling (ABM) engine with a focus on epidemiological models. The core of the tool has been implemented in C++--available at <https://github.com/UofUEpiBio/epiworld>--and it has [R](https://github.com/UofUEpiBio/epiworldR) and [Python](https://github.com/UofUEpiBio/epiworldpy) wrappers.

## Architecture

Althought `epiworld` contains a large number of modules and moving parts, the most important ones are the `Model`, `DataBase`, `Queue`, `Event`, `Agent`, `Virus`, `Tool`, and `Entity`.

```{mermaid}
%%| fig-cap: "Main components of `epiworld` and how they are interrelated."
flowchart TD
  subgraph SE[Simulation Engine]
    M[Model]
    D[DataBase]
    Q[Queue]
    E[Events]
  end

  M -->|Reads/Writes| D
  M -->|Optimizes via| Q
  M -->|Triggers| E

  subgraph In[Instances]
    A[Agent]
    V[Virus]
    T[Tool]
    En[Entity]
  end

  A -.->|"Belongs to<br>(optional)"| En
  A -->|Carries| V
  A -->|Uses| T


  M -->|Coordinates| A
  E -->|Modifies| A
```

From the figure, we can hihghlight a few things. First, all classes instances live within the `Model` class; `Agents` or `Virus`es outside of the model have no meaning for `epiworld`. Second, the `DataBase` class keeps track of all the revelant information, including **daily counts**, **trasmission events**, **viruses** and **tools**, etc. Third, although `Agent`s can carry a single virus at a time, models can feature multiple viruses simulatenosly[^why-single-virus], models can feature multiple viruses in the same run, *e.g.*, multipe strands of COVID-19. Fourth, agents can feature multiple `Tool`s: a class of object that alters agents' susceptibility, infectiosness, and recovery, *e.g.*, a vaccine.

[^why-single-virus]: At some point in time, `epiworld` supported agents beign infected with multiple viruses at once. Nonetheless, although an interesting feature, it was computationally expensive. Thus, we decided to simplify the tool and feature a single virus per agent.

The library features two types of contact dynamics: network-based interaction and mixing models. In the network based interaction--the fastest models in `epiworld`--transmission takes place through a pre-defined network. In the case of mixing models, agents are organized in `Entity`(ies) which determines group-level interaction. Mixing models bridge ABMs with compartmental models, providing an approximation that does not require specifying a network structure.

::: {.callout-tip title="Network vs Mixing models"}
Although network-based models are faster, they usually assume a known contact network which may not be available. Mixing models provide a nice approximation that essentially integrates-out the uncertainty about network-based models. In principle, the same could be done with network-based models, but, in order to do so, we would need to embed the graph-generating algorithm in the model, which could increase the computational complexity.
:::

## Dynamics

`epiworld` is a discrete-time ABM, meaning that, instead of handling a queue of updates and events, it collects updates from a single step and applies them at the end of it. For the user, this simply means that changes in the system that were triggered in time $t$ will be reflected in time $t + 1$. Each run of epiworld is characterized by the following sequence:

1. **Reset** the model to its initial state: clear the database, reset the agents, and reset the entities.

2. **Distribute** agents to entities, viruses to agents, and tools to agents. Each call finalizes with `run_events()`.

::: {.callout-note title="Locking changes"}
The functions `set_virus()`, `add_tool()`, `add_entity()` or `update_state()` are not effective until `run_events()` is called. Since epiworld is discrete-time, changes must be applied simultaneously.
:::

3. **Main loop**. For each day in `1 ... ndays` do:

    a. **Queue**: Check if agents is in the queue. If not, skip to the next iteration.
    b. **Update Agents**If the agent is in the queue, trigger the update function associated to the agent, *e.g.*, `update_susceptible()` if the agent is susceptible, `update_infected()` if the agent is infected, etc.
    c. **Global Events**: Iterate FIFO through global events. This could trigger more updates to agents, viruses, tools, and entities. For example, a global event could be a lockdown that reduces the contact rate of all agents by 50% for 14 days.
    d. **Save changes**: Save the state of the model in the database and increment the day counter.

4. **Return**: The simulation ends and the user can retrieve information from it.


Because `epiworld` has been designed to be fast, we generally run multiple simulations, generating a distribution of outcomes instead of single point estimates.

## Example: with SEIR model

In this example, we will create a simple SEIR model using a Stochastic-Block Model for generating a random network (included in `epiworldR`). This creates a network with, in this case, two groups of 500 agents each, with a higher contact rate within groups than between groups. 

### Running the simulation

```{r}
#| label: example-1make 
# Initializing SIR model with SBM network
library(epiworldR)
sir <- ModelSIR(
  name = "COVID-19",
  prevalence = 5/1000,
  transmission_rate = 0.01,
  recovery_rate = 1/7
)

# Creating the SBM network featuring two groups
agents_sbm(
  sir,
  block_sizes = c(500, 500),
  mixing_matrix = matrix(c(20, 5, 5, 20), nrow = 2)
)

# Running the simulation
run(sir, ndays = 365, seed = 1912)
```

After the simulation takes place, we can use the methods `summary()` and `plot()` to get a first look at the results:

```{r}
#| label: summary
summary(sir)
```

The `summary()` method prints the following information:

- The model metadata, including, name, population, duration, and elapsed time.
- Information about events (which we don't have yet), viruses (single), and tools (none).
- Model parameters (which can be modified via `set_param()`).
- Distribution of the agents from time 0 to `ndays` (365 in this case), and
- The transition matrix, where each entry indicates the trasition probability from one state to another.

The `plot()` method provides a simple view of the outbreak too

```{r}
#| label: simple-plot
plot(sir)
```

But we can get much more detailed information! Mostly using the following functions:

- `get_active_cases()`
- `get_generation_time()`
- `get_hist_total()`
- `get_hist_transition_matrix()`
- `get_hospitalizations()`
- `get_outbreak_size()`
- `get_reproductive_number()`
- `get_transition_probability()`
- `get_transmissions()`

Although, most names are self-explainatory, we will try them out later.

### Visualizing the results

```{r}
#| label: incidence-and-rt
op <- par(mfrow = c(1, 2))
plot_reproductive_number(sir)
plot_incidence(sir)
par(op)
```

We can even generate the transmission network (which should be clustered). We use the `data.table` R package for data wrangling and `netplot` for network visualization: 

```{r}
#| label: network-vis
#| message: false
#| warning: false
library(data.table)
library(netplot)
library(igraph)

# Extracting the transmission tree
net <- get_transmissions(sir) |>
  data.table()

# Visualizing the transmission tree
net <- graph_from_edgelist(
  el = as.matrix(net[source != -1, .(source, target)] + 1),
  directed = TRUE
)

set.seed(442)
nplot(
  net,
  edge.line.breaks = 1,
  layout           = layout_as_tree(net),
  skip.arrows      = TRUE,
  skip.vertex      = TRUE,
  edge.color       = "darkgray"
  )
```

## Exercises

1. *Basic model*. Look for the `ModelSEIRCONN()` and create an Agent-Based model featuring 10,000 agents. Run a simulation with the following characteristics: (a) has a virus with 1 initial case, (b) has a transmission rate of 0.1 with a reproductive number of 2 (so you need to figure out the contact rate), (c) runs for 50 days.

::: {.callout-tip title="Simple ABM calibration"}
In the case of a simple ABM (based on a fixed contact and transmission rate), we can use the following equation to calibrate our models:

$$
\mathcal{R}_0 = p_t \times c / p_r \implies c = \mathcal{R}_0 \times p_r / p_t
$$

Where $c$ is the contact rate, $p_r$ is the recovery rate, and $p_t$ is the transmission rate.
:::

2. *Multiple runs*. Look at the documentation of `run_multiple()` and (a) create a saver that extracts the outbreak size, (b) execute the model for the same 50 days, and (c) exctract the saved outbreak size and plot the distribution.

3. *A network model*. Using the `ModelSEIR()` function, (a) create a version of the model using a network instead (you can use the small-world network), (b) replicate the `run_multiple()` exercise. and (d) compare the distribution with and without network

::: {.callout-tip title="Calibration of Network Models"}
Calibrating network-based models is not a simple task. The traditional sense of the basic reproductive number is somewhat misleading in network-based models overestimating the quantity (see @huangEndogenousCompetitionUnderrealized2025). Nonetheless, for the moment, it will still give you a good starting point.
:::

## References
