---
title: "SEIR outbreak simulation with mixing and quarantine in Miami"
---

::: {.callout-note}
This example has been adapted from the ongoing work on the FIFA World Cup, InsightNet's ID Cup <https://github.com/EpiForeSITE/idcup>.
:::

Now, we will build a more complex model using the `ModelSEIRMixingQuarantine()` model. The new model has the following components:

- *Contact Matrix*: Using data from Epistorm [@litvinovaEpistormMixMappingSocial2025], this provides a contact matrix at the group level, indicating the average number of contacts between age groups.
- *Heterogenous Vaccination*: We can specify vaccination rates to be heterogenous at the group level.
- *Contact tracing and isolation*: This is built-in feature in the model. Although users could implement their own using R, it is not recommended as the C++ version is highly efficient.



## Preparing the environment

```{r}
#| label: loading-libraries
#| warning: false
#| message: false
library(epiworldR)
library(measles)
library(data.table)
library(ggplot2)

params <- list(
  city     = "Miami",
  max_pop  = 50000,
  nthreads = 4,
  seed     = 8812,
  ndays    = 60,
  nsims    = 100
)
```

Pulling information about the city and population from the data files. The mixing matrix is based on the EpiStorm [@litvinovaEpistormMixMappingSocial2025] survey data, which provides age-based contact patterns for various cities. We will use the mixing matrix corresponding to the specified city in `params$city`.

```{r}
#| label: loading-data
# Population matrix
census_age <- fread("data/census_age.csv")[city == params$city]
population <- 487014

# Mixing matrix (age-based, from Epistorm survey data)
mixing_matrices <- readRDS("data/mixing_matrix.rds")
```

The total population in the census data `r prettyNum(population)`, exceeds the maximum population size (`r prettyNum(params$max_pop)`). The population will be scaled down to fit within the specified maximum population size.

```{r}
#| label: adjusting-population
N <- min(params$max_pop, sum(population))

census_age[, total := ceiling(total / sum(total) * N)]
N <- sum(census_age$total)
```

We calibrate the contact matrix to a target basic reproduction number of 2.5, consistent with a moderately transmissible respiratory pathogen (e.g., pandemic influenza). The infectious period is set to 7 days and the per-contact transmission probability to 0.1.

```{r}
#| label: calibrating-mixing
calib_scale <- calibrate_mixing_model(
  contact_matrix         = mixing_matrices,
  target_rep_number      = 2.5,
  infectious_period_days = 7,
  transmission_prob      = 0.1
)
```

## Baseline scenario

```{r}
#| label: creating-model

# Hospitalization rate such that P(hosp) = 20%
# P(hosp) = hosp_r / (hosp_r + rec_r)
#   => hosp_r = P(hosp) * rec_r / (1 - P(hosp))
h_rate <- 0.2 * (1 / 7) / (1 - 0.2)

seir_model <- ModelSEIRMixingQuarantine(
  name                         = "Respiratory pathogen",
  n                            = N,
  prevalence                   = 1 / N,
  transmission_rate            = 0.1,
  incubation_days              = 5,
  recovery_rate                = 1 / 7,
  contact_matrix               = mixing_matrices * calib_scale,
  hospitalization_rate         = h_rate,
  hospitalization_period       = 7,
  # Detection and intervention parameters
  days_undetected              = 3,
  quarantine_period            = 14,
  quarantine_willingness       = 0.8,
  isolation_willingness        = 0.6,
  isolation_period             = 7,
  # Contact tracing parameters
  contact_tracing_success_rate = 0.7,
  contact_tracing_days_prior   = 4
)
```

```{r}
#| label: adding-entities
# Adding age-group entities to the model
add_entities_from_dataframe(
  model        = seir_model,
  entities     = census_age,
  col_name     = "age_group",
  col_number   = "total",
  as_proportion = FALSE
)
```

We will use a single initial infectious case seeded randomly in the population.

```{r}
#| label: distribute-initial-cases
get_virus(seir_model, 0L) |>
  set_distribution_virus(
    distfun = distribute_virus_randomly(
      prevalence    = 1,
      as_proportion = FALSE
    )
  )
```

We can now run the simulations.

```{r}
#| label: running-simulations
saver <- make_saver("outbreak_size", "hospitalizations", "transition")

seir_model |>
  run_multiple(
    ndays    = params$ndays,
    nsims    = params$nsims,
    seed     = params$seed,
    saver    = saver,
    nthreads = params$nthreads
  )
```

We can check the results of the simulations by summarizing the model:

```{r}
#| label: checking-simulations
summary(seir_model)
```

## Extracting and plotting the results

```{r}
#| label: extracting-results
ans <- seir_model |>
  run_multiple_get_results(
    freader = data.table::fread
  )

outbreak_size    <- ans$outbreak_size    |> as.data.table()
hospitalizations <- ans$hospitalizations |> as.data.table()
transition       <- ans$transition       |> as.data.table()
```

### Transition

```{r}
#| label: plotting-transition
tmat <- get_transition_probability(seir_model) * 0

t_totals <- transition[, .(total = sum(counts)), by = .(from, to)]
tmat[as.matrix(t_totals[, .(from, to)])] <- t_totals$total

draw_mermaid_from_matrix(tmat) |>
  plot()
```

### Outbreak size

```{r}
#| label: plotting-outbreak-size1
ggplot(
  outbreak_size,
  aes(x = date, y = outbreak_size)
) +
  geom_point(
    color = "steelblue",
    alpha = 0.2,
    size  = 1.5
  ) +
  labs(
    title    = paste("SEIR outbreak size in", params$city),
    subtitle = sprintf(
      "Mixing model with census data (%i simulations)",
      params$nsims
    ),
    x = "Day",
    y = "Cases"
  ) +
  theme_minimal()
```

```{r}
#| label: plotting-outbreak-size2
final_outbreak      <- outbreak_size[date == max(date)]
median_outbreak_size <- median(final_outbreak$outbreak_size)

ggplot(final_outbreak, aes(x = outbreak_size)) +
  geom_histogram(fill = "steelblue", color = "white") +
  labs(
    title    = sprintf(
      "SEIR outbreak size in %s (median: %.0f)",
      params$city,
      median_outbreak_size
    ),
    subtitle = sprintf(
      "Mixing model with census data (%i simulations)",
      params$nsims
    ),
    x = "Final outbreak size",
    y = "Count"
  ) +
  theme_minimal()
```

### Hospitalizations

```{r}
#| label: plotting-hospitalizations
hosp_daily <- hospitalizations[, .(Total = sum(count)), by = .(sim_num)]

ggplot(hosp_daily, aes(x = Total)) +
  geom_histogram() +
  labs(
    title    = paste(
      "Daily hospitalizations in SEIR outbreak in",
      params$city
    ),
    subtitle = sprintf(
      "SEIR mixing model (%i simulations)",
      params$nsims
    ),
    y = "Frequency",
    x = "Total Hospitalizations"
  ) +
  theme_minimal()
```


## Session info

This document was generated using the `{epiworldR}` package version `r packageVersion("epiworldR")` on `r Sys.Date()`. We used R version `r getRversion()`, running on `r Sys.info()[['sysname']]` with `r Sys.info()[['machine']]` architecture.

## References
