---
title: SEIR model
---

In this example, we will create a simple SEIR model using a Stochastic-Block Model for generating a random network (included in `epiworldR`). This creates a network with, in this case, two groups of 500 agents each, with a higher contact rate within groups than between groups. 

### Running the simulation

```{r}
#| label: example-1make 
# Initializing SIR model with SBM network
library(epiworldR)
sir <- ModelSIR(
  name = "COVID-19",
  prevalence = 5/1000,
  transmission_rate = 0.01,
  recovery_rate = 1/7
)

# Creating the SBM network featuring two groups
agents_sbm(
  sir,
  block_sizes = c(500, 500),
  mixing_matrix = matrix(c(20, 5, 5, 20), nrow = 2)
)

# Running the simulation
run(sir, ndays = 365, seed = 1912)
```

After the simulation takes place, we can use the methods `summary()` and `plot()` to get a first look at the results:

```{r}
#| label: summary
summary(sir)
```

The `summary()` method prints the following information:

- The model metadata, including, name, population, duration, and elapsed time.
- Information about events (which we don't have yet), viruses (single), and tools (none).
- Model parameters (which can be modified via `set_param()`).
- Distribution of the agents from time 0 to `ndays` (365 in this case), and
- The transition matrix, where each entry indicates the trasition probability from one state to another.

The `plot()` method provides a simple view of the outbreak too

```{r}
#| label: simple-plot
plot(sir)
```

But we can get much more detailed information! Mostly using the following functions:

- `get_active_cases()`
- `get_generation_time()`
- `get_hist_total()`
- `get_hist_transition_matrix()`
- `get_hospitalizations()`
- `get_outbreak_size()`
- `get_reproductive_number()`
- `get_transition_probability()`
- `get_transmissions()`

Although, most names are self-explainatory, we will try them out later.

### Visualizing the results

```{r}
#| label: incidence-and-rt
op <- par(mfrow = c(1, 2))
plot_reproductive_number(sir)
plot_incidence(sir)
par(op)
```

We can even generate the transmission network (which should be clustered). We use the `data.table` R package for data wrangling and `netplot` for network visualization: 

```{r}
#| label: network-vis
#| message: false
#| warning: false
library(data.table)
library(netplot)
library(igraph)

# Extracting the transmission tree
net <- get_transmissions(sir) |>
  data.table()

# Visualizing the transmission tree
net <- graph_from_edgelist(
  el = as.matrix(net[source != -1, .(source, target)] + 1),
  directed = TRUE
)

set.seed(442)
nplot(
  net,
  edge.line.breaks = 1,
  layout           = layout_as_tree(net),
  skip.arrows      = TRUE,
  skip.vertex      = TRUE,
  edge.color       = "darkgray"
  )
```

## Exercises

1. *Basic model*. Look for the `ModelSEIRCONN()` and create an Agent-Based model featuring 10,000 agents. Run a simulation with the following characteristics: (a) has a virus with 1 initial case, (b) has a transmission rate of 0.1 with a reproductive number of 2 (so you need to figure out the contact rate), (c) runs for 50 days.

::: {.callout-tip title="Simple ABM calibration"}
In the case of a simple ABM (based on a fixed contact and transmission rate), we can use the following equation to calibrate our models:

$$
\mathcal{R}_0 = p_t \times c / p_r \implies c = \mathcal{R}_0 \times p_r / p_t
$$

Where $c$ is the contact rate, $p_r$ is the recovery rate, and $p_t$ is the transmission rate.
:::

2. *Multiple runs*. Look at the documentation of `run_multiple()` and (a) create a saver that extracts the outbreak size, (b) execute the model for the same 50 days, and (c) exctract the saved outbreak size and plot the distribution.

3. *A network model*. Using the `ModelSEIR()` function, (a) create a version of the model using a network instead (you can use the small-world network), (b) replicate the `run_multiple()` exercise. and (d) compare the distribution with and without network

::: {.callout-tip title="Calibration of Network Models"}
Calibrating network-based models is not a simple task. The traditional sense of the basic reproductive number is somewhat misleading in network-based models overestimating the quantity (see @huangEndogenousCompetitionUnderrealized2025). Nonetheless, for the moment, it will still give you a good starting point.
:::

## References
